---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# DEPM

<!-- badges: start -->
<!-- badges: end -->

## Installation

You can install the released version of DEPM from [Github](https://github.com/jonathansmart/DEPM) with:


```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("jonathansmart/DEPM")
```


The DEPM package is a tool to estimate Spawning Stock Biomass for fish populations using Daily Egg Production Methods (DEPM). Two methods are available in this package:

1. DEPM - the traditional DEPM which uses mean weight and mean fecundity to scale the number of females to total biomass
2. DEPMWt - a new approach to DEPM created in McGarvey et al. (in review), where female weigth and fecundity are grouped in weight bins. This approach is useful for demersal finfish whose weight distributions are often not unimodal and therefore should be described using a mean.

The package contains functions that estimate the individual parameters for DEPMs: daily egg production ($P_{0}$), sex ratio (*R*), spawning fraction (*S*), fecundity (both mean and weight specific) and weight (both mean and in weight bins). The only parameter that is not estimated in this package is spawning area (*A*), which is typically estimated using GIS methods and must be provided to the analysis.

# How to guides 

Two vignettes are available that wil guide users through both approaches and can be accessed using:

```{r, eval=FALSE}
browseVignettes("AquaticLifeHistory")
```

