---
title: "DEPM_Estimation"
author: "Jonathan Smart"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
# output: word_document
vignette: >
  %\VignetteIndexEntry{DEPM Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load the DEPM and tidyverse libraries.
```{r setup, warning=FALSE, message=FALSE}
library(DEPM)
library(tidyverse)
```

# Introduction

The traditional Daily Egg Production Method (DEPM) has been used to determine spawning stock biomass of fish populations for over 40 years. While this package is primarily designed to facilitate the use of revised approaches to DEPM, including  (stage-based estimation of daily egg production (McGarvey et al. 2018) and female adult size-dependence of population numbers and batch fecundity (DEPMWt; McGarvey et al in review), many aspects of the approach (i.e. estimation of some key DEPM parameters) remain the same. Therefore, the traditional DEPM methods are also included in this package.  Here, an example of the DEPM approach is given using South Australian snapper (*Pagrus auratus*).

# Necessary data
Five datasets are required to use the DEPM approach:

* Egg density data to estimate daily egg production ($P_{0}$)
* Adult sex ratios to estimate *R*
* Female spawning fractions to estimate *S*
* Female total weights to estimate mean adult female weight (*W*)
* Female total weight and batch fecundities to estimate mean female fecundity (*F*)

Spawning Area (*A*) is not calculated as part of this package but its estimates will be included as one of the final steps.

```{r, include = FALSE}

library(scales)
library(lubridate)
# Data for P0
egg_data_raw <- read.csv("G:\\ModelGrp\\Snapper\\DEPM 2019 method comparison\\Data\\raw_egg_density_Snapper_18.csv")


## Ricks functions to determine Age~Stage based on transforming Mikes 9 stages to McGlennons 16 stages
Age.sMcGlenT=function(sMcGlen,TA){
  age=36.158 * exp(-0.12*TA) * (sMcGlen)^0.827
  return(age/24)
}

Age.sSteerT=function(sSteer,TA) {
  sMcGlenFromsSteer=c(1.5,3.5,5.5,7.5,9,10.5,12.5,14.5,16)
  AgeInDays=Age.sMcGlenT(sMcGlenFromsSteer[sSteer],TA)
  return(AgeInDays)
}

# Determine hatching time
HatchingTime.T=function(TH){  #test TH=20.64252
  H = Age.sMcGlenT(16,TH) + (Age.sMcGlenT(16,TH) - Age.sMcGlenT(15,TH))/2
  return(H)
}


egg_data_raw <- egg_data_raw %>% mutate(SST = ifelse(is.na(SST), lag(SST), SST),
                                SST = ifelse(is.na(SST), lag(SST), SST),
                                SST = ifelse(is.na(SST), lag(SST), SST))


# calculate Density based on the surface area of the net (multiplied by 2 within excel data to account for two
# nets being used), mean flowmeter distance (each net has one) and the depth determined by the CTD. This
# is the necessary formula for oblique tows and therefore differs to vertical tows in previous surveys.
stage_data <- egg_data_raw %>% mutate(
  Density_I = (1/Surface.area.of.Net..m2.)*(1/mean.flowmeter.distance)* Stage_I*Actual.Depth,
  Density_II = (1/Surface.area.of.Net..m2.)*(1/mean.flowmeter.distance)*Stage_II*Actual.Depth,
  Density_III = (1/Surface.area.of.Net..m2.)*(1/mean.flowmeter.distance)*Stage_III*Actual.Depth,
  Density_IV = (1/Surface.area.of.Net..m2.)*(1/mean.flowmeter.distance)*Stage_IV*Actual.Depth,
  Density_V = (1/Surface.area.of.Net..m2.)*(1/mean.flowmeter.distance)*Stage_V*Actual.Depth,
  Density_VI = (1/Surface.area.of.Net..m2.)*(1/mean.flowmeter.distance)*Stage_VI*Actual.Depth,
  Density_VII = (1/Surface.area.of.Net..m2.)*(1/mean.flowmeter.distance)*Stage_VII*Actual.Depth,
  Density_VIII = (1/Surface.area.of.Net..m2.)*(1/mean.flowmeter.distance)*Stage_VIII*Actual.Depth,
  Density_IX = (1/Surface.area.of.Net..m2.)*(1/mean.flowmeter.distance)*Stage_IX*Actual.Depth)


egg_data <- stage_data %>% dplyr::select(Gulf,Region,Station.ID.code, SST, starts_with("Density")) %>%
  gather(Stage, Density, -Gulf, -Region, -Station.ID.code, -SST) %>%   
  mutate(Stage_num = as.numeric(as.roman(sub("Density_","", x =  Stage))),
         Age = Age.sSteerT(Stage_num,SST),
         Hatch = HatchingTime.T(SST)) %>% 
  select(Region, Site = Station.ID.code,Stage_num, Density, Age, Hatch)

# Data for S (Used for all surveys)
S_data <- read.csv("G:\\ModelGrp\\Snapper\\DEPM 2019 method comparison\\Data\\Spawning_fraction.csv") %>% rename(yes = Females) %>% 
  mutate(no = Total-yes) %>% .[c(1,3,2)]

# Data for R (2018 only)
R_data <- read.csv("G:\\ModelGrp\\Snapper\\DEPM 2019 method comparison\\Data\\Sex_ratio.csv") %>% 
  rename(Tot.Wt = Tot_WT..kgs.) %>%
  mutate(Year = lubridate::year(lubridate::dmy(Processing_date)),
         Sample.code = paste0(Sample.code, Processing_date)) %>%
  filter(Sex %in% c("M", "F")) %>%
  group_by(Year, Region,Sample.code, Sex) %>%
  summarise(Wt = sum(Tot.Wt , na.rm = T)) %>%
  spread(Sex, Wt) %>%
  mutate(`F` = ifelse(is.na(`F`), 0, `F`),
         M = ifelse(is.na(`M`), 0, `M`),
         Total = `F`+ M) %>% 
  select(-Sample.code) %>% as.data.frame()

# Data for Adult Wt (All years)
Wt_data <- read.csv("G:\\ModelGrp\\Snapper\\DEPM 2019 method comparison\\Data\\TotWt_GFwt.csv")  %>%
  mutate(Tot.Wt = Tot_WT..kgs.*1000, Gonad_free_Wt = Tot.Wt - Gonad.Wt) %>%
  filter(Sex == "F", Gonad.Wt < Tot.Wt, Year == 2018)

# Data for Batch Fecundity
Batch_Fec_data <- read.csv("G:\\ModelGrp\\Snapper\\DEPM 2019 method comparison\\Data\\batch_fecundity_data.csv")

# raw fecundity data for total weight
fecundity_data_GonardFrWT <- Batch_Fec_data %>%
  select(Gonad_fr_wt,batch_fecund )

```

# DEPMWt parameter estimation
## Daily egg production ($P_{0}$)

Daily egg production is estimated using a stage-based estimator (McGarvey et al. 2018) via the function `Estimate_P0()` which requires a dataset that has been formatted as such:

```{r}
# egg density structure
str(egg_data)

# egg density head()
head(egg_data)

```

This dataset is in a long format where every row is a density estimate of the eggs in each stage of each sample. This dataset contains several columns that must include:

* Site - which represents each sample in the data (i.e. one plankton tow a location). It's variable name must be specified as an argument using `site = `
* Density - the number of eggs in a single development stage in a metre cubed of seawater. The function will automatically determine the density column based on similar names (i.e. dens, density, DENSITY or other similar names). However, there cannot be multiple density columns.
* Age - the age in days of each egg stage in each sample. The function will automatically determine the Age column based on similar names in a similar manner to the `Density` variable
* Hatching time - The age in days where the eggs in each sample are estimated to hatch (usually differs based on temperature). The function will automatically determine the `Hatch` column based on similar names in a similar manner to the `Density` variable
* Z - a pre-specified level of egg mortality used to determine $P_{0}$. Multiple values can be provided to simultaneously determine different values of $P_{0}$ based on this parameter. Instruction on how to do this and how to perform sensitivity tests are provided in the final section of this vignette.

Additional columns can be included in the dataset, so the user does not need to remove variables that could be useful to them later on. For example, in the `egg_data` dataset, the `Stage_num` variable represents the development stage of each density estimate in each sample. However as each row of this dataset is a different observation, this variable is not needed by the function.

The function also has the ability to break the data down spatially and temporally. If the `Region` and `Time` arguments are not used then $P_{0}$ will be estimated using all of the data. If you provide either `Region` or `Time` then a $P_{0}$ will be returned for each Region/Time combination. Therefore, a time series of data can analysed in a single function call. Here is an example using 4 DEPM surveys conducted in different areas but in the same year.
The results are returned as a dataframe where each row is a survey and the estimate, standard error and specified mortality are returned.

```{r}
# run P0 function
P0_results <- Estimate_P0(data = egg_data,
                          site = "Site", 
                          Region = "Region", 
                          Z = .4)

P0_results
```

## Spawning fraction (*S*)

Spawning fraction is estimated using the function `Estimate_Spawning_fraction()` which requires a dataset that has been formatted as two columns: 1) the number of females in spawning condition and 2)the total number of females. Each row represents a sample and the spawning fraction is estimated using a ratio estimator.

The correct columns in the dataset will be detected by the function. The column with the total number of females should be called "Total", "Tot" or something similar. The column with the number of spawning females can include the term "spawn" or "yes" depending on how your dataset is setup. Ours is named "yes" as we had a "yes" or "no" designation for spawning fish.



```{r}

str(S_data)

head(S_data)


```

Similar to other functions, specifying the `Region` or `Time` argument with relevant column names will group the estimates according to specific surveys. However, here an example is given without a region or time grouping. 
The results are returned as a dataframe with the estimate, its variance, the standard error and its CV. Each row represents a survey.


```{r}
Spawn_results <- Estimate_Spawning_fraction(S_data)

Spawn_results
```


## Sex Ratio (R)

Sex ratio is estimated using the function `Estimate_sex_ratio()` which requires a dataset that has been formatted to include the weight of males and females along with `Region`/`Time` variables. Each row represents a sample and the sex ratio is estimated using a ratio estimator in the same manner as spawning fraction. The same Region/Time break down can be applied and Region is available in this dataset. The results returned are the same format as `Estimate_Spawning_fraction()`.

```{r}

head(R_data)

sex_ratio_results <- Estimate_sex_ratio(R_data, Region = "Region")

sex_ratio_results

```

## Female weight  and fecundity 

The key difference between teh DEPMWt approach and the traditional DEPM is how female weight and fecundity are handled. In the DEPMWt approach, female weight is broken into discrete weight bins and their variance is described using a multinomial distribution. However, in the traditional DEPM 
approach, the *W* parameter is used and represents the mean female weight and its variance. Similarly, fecundity is also expressed as the mean number of eggs per adult female and its variance.

**Fecundity**

Fecundity needs to be summarised in a couple of ways. Firstly, a batch-fecundity relationship is required for **gonad free weight** which can be used to predict the fecundity of fish whose histology has not been examined. This relationship can be estimated using the `Estimate_Batch_Fecundity()` function. This function is quite useful as it can return three sets of outputs:

* A data.frame of parameters and their variances for the relationship (these can also be printed to the screen) by setting `return.parameters = TRUE`.
* A dataframe with the predicted fecundity for each weight along with SD and 95% confidence intervals
* A dataframe with the predicted fecundity at specified weights and their variance.

The batch fecundity estimator uses an allometric relationship which allows for wider variance with larger weights. Therefore, there are four parameters returned, alpha, beta and two sigma parameters that determine how variance changes with weight.
A dataframe with two columns must be provided that includes **gonad free weight in grams** and the number of eggs for that fish. The function will determine which is which based on their scales (number of eggs > Wt in grams). A set of starting parameters are required and are provided to the `start_pars` argument as a list. The estimated parameters are printed to the screen if `verbose = TRUE`.


```{r fig.height=6, fig.width=6}

head(fecundity_data_GonardFrWT)

# List of starting parameters for the batch fecundity estimator
parameters <- list( alpha= 110, beta = 0.85 , Sigma0 = 10.5, Sigma1 = 0.7)

# Return parameters
Batch_fec_pars <- Estimate_Batch_Fecundity(fecundity_data_GonardFrWT, start_pars = parameters, return.parameters = TRUE)

# Return batch fecundity estimates
Batch_fecundity_relationship <- Estimate_Batch_Fecundity(fecundity_data_GonardFrWT, start_pars = parameters, verbose = FALSE)

#plot of relationship
ggplot(Batch_fecundity_relationship, aes(Wt, y = Predicted, ymin = low, ymax  = upp))+
  geom_ribbon( alpha = .3)+
  scale_y_continuous(labels = scales::comma)+
  geom_point(aes(y = Fecundity), alpha = .3)+
  geom_line(aes(y = Predicted), col = "royalblue", size = 1) +
  theme_bw()

```

**Weight**
The parameters for *W* and *F* are determined using a single function called `Estimate_mean_W_F()`. This function requires:

* A dataframe of `weight.data` which will be used to estimate *W* and estimate *F* based on the fecundity-at-weight relationship.
* A character string for `TotalWt` indicating the variable name for Total weight in grams. This will be used to estimate *W*.
* A character string for `GonadFrWt` indicating the variable name for gonad free weight in grams. This will be used to estimate *F* for each weight. If gonad free weight is not available, then this argument can be left `NULL` which will prompt the function to estimate fecundity based on total weight instead. **NOTE: If total weight is used to determine *F* then total weight should be used in return batch fecundity parameters from  `Estimate_Batch_Fecundity()` as well  **
* A dataframe of batch fecundity relationship parameters from `Estimate_Batch_Fecundity()` with `return.parameters = TRUE`. The object returned from this call can be directly provided to `Estimate_mean_W_F()`.

As with other functions, providng `Time`/`Region` variable names will group the results either spationally, temporally or both. Here a `Region` grouping is applied.

```{r}
# data structure
head(Wt_data)

# Estimate mean W and mean F using total weight and gonad free weight, respectively.
F_W_results <- Estimate_mean_W_F(weight.data = Wt_data,
                              TotalWt = "Tot.Wt",
                              GonadFrWt = "Gonad_free_Wt",
                              Region = "Region",
                              parameters = Batch_fec_pars)
F_W_results
```

## Spawning Area (*A*)

Spawning area is the only DEPM parameter that is not estimated in this package as this is typically done using GIS methods. However, the Spawning area object can be created manually by taking another parameter object (*P0, S* or *R* are good choices) and using its `Time`/`Region` groupings to create new object. As this example has one survey from 4 regions in a single year, only the `Region` variable needs to be carried across. A new variable can then be added with the location in the correct order in that column. Spawning Area is a precise quantity and therefore does not require a variance. Note that spawning area must be in metres squared.



```{r}
# Get Time/region combos from the P0 object
Area <- select(P0_results, Region)

Area$A <- c(2822201039, #NGSV
            1884895030, # NSG
            2285649237, # SGSV
            3239913469) # SSG

head(Area)
```



# Combining parameters as inputs for biomass estimation

With all of the parameters now available to calculate spawning stock biomass, these now need to be combined into a dataset that can be input into the `Estimate_DEPM_Biomass()` function. For the DEPM approach, two objects are needed:

## Combining parameter estimates
A dataframe of all of the parameter estimates can be assembled using `combine_estimates()` where each of the parameter objects are provided as arguments.


```{r}
Adults_pars <- combine_estimates(P0 = P0_results, R = sex_ratio_results, S = Spawn_results, A = Area, W_F = F_W_results)
Adults_pars
```

Where a parameter is estimated as time invariant, it will be automatically allocated to all of the `Region` or `Time` groupings applied for the other parameters. Here this occurs for *S*. However, if a particular parameter is missing in one instance but is not time invariant, it will not be carried over and `NA` will be returned. In this situation, it will need to be inserted manually. For example, lets pretend that *R* is missing for "SSG":

```{r}
Adults_pars[Adults_pars$Region == "SSG", "R"] <- 0.5741385
```

## Combining parameter variances

Variance estimates are needed by `Estimate_DEPM_Biomass()` in order to determine the precision of the biomass estimates. These are provided in the same fashion as the parameter estimates using `combine_variances()`.

```{r}
Adults_vars <- combine_variances(P0_results, R = sex_ratio_results, S = Spawn_results, W_F = F_W_results)
Adults_vars
```

The same manual insertions can be performed on this object:

```{r}
Adults_vars[Adults_vars$Region == "SSG", "R"] <- 0.0018484630
```

# Spawning Stock Biomass Estimation

All of the hard work is done once all of the parameters and their variances have been estimated and organised. The final step is to combine them using `Estimate_DEPM_Biomass()`. This function requires each of these objects and will return a list of results which include the spawning stock biomass and total number of females.

```{r}
# Estimate Biomass
Snapper_biomass_results <- Estimate_DEPM_Biomass(adult.pars = Adults_pars,
                                          adult.vars = Adults_vars)
# a list of two outputs is returned
Snapper_biomass_results$Biomass # biomass in kgs  
Snapper_biomass_results$Nfem # number of females

```

The list of dataframes returned from the function can then be saved and used for plotting

```{r}

# Create new objects
Snapper_biomass <- Snapper_biomass_results$Biomass
Snapper_Nfem <- Snapper_biomass_results$Nfem

```

**Plot biomass with SD**
```{r fig.height=6, fig.width=6}

ggplot(Snapper_biomass, aes(Region,Biomass, 
                            ymin = Biomass - SD,
                            ymax = Biomass + SD))+
  geom_crossbar(fill = "royalblue", alpha = .8)+
  expand_limits(y = 0)+
  scale_y_continuous(name = "Biomass (kg)", labels = scales::comma, breaks = scales::pretty_breaks(5))+
  theme_bw()

```

**Plot total number of females**
```{r fig.height=6, fig.width=6}

ggplot(Snapper_Nfem, aes(Region,Nfem, 
                            ymin = Nfem - SD,
                            ymax = Nfem + SD))+
  geom_crossbar(fill = "purple", alpha = .8)+
  expand_limits(y = 0)+
  scale_y_continuous(name = "Number of females", labels = scales::comma, breaks = scales::pretty_breaks(5))+
  theme_bw()

```

# Testing assumptions for pre-specified egg mortality (*Z*)

As the $P_{0}$ estimation methods in this package require a specified level of egg mortality (*Z*), it is sensible to test the sensitivities of the $P_{0}$ estimates, as well as the DEPMWt results to different *Z* values. Testing different values of *Z* can be done by passing a vector of values to the `z` argument of `Estimate_P0()`:

```{r}
Multi_Z_P0_results <- Estimate_P0(data = egg_data,
                          site = "Site", 
                          Region = "Region", 
                          Z = c(.2, .4, .6))

```

This will return a different $P_{0}$ estimate for each value of *Z* for each `Time`/`Region` grouping:

```{r}
Multi_Z_P0_results
```

For South Australian Snapper, the resulting $P_{0}$ estimates are very similar.

```{r fig.height=6, fig.width=6}

Multi_Z_P0_results$Z <- as.factor(Multi_Z_P0_results$Z) # convert Z to a factor for easy plotting

ggplot(Multi_Z_P0_results, aes(Z, P0, ymin = P0 - P0_se, ymax = P0 + P0_se, fill = Z))+
  geom_crossbar()+
  facet_wrap(~Region, ncol = 2, scales = "free")+
  scale_fill_viridis_d()+
  theme_bw()

```

An object from `Estimate_P0()` with multiple *Z*s can be included in the biomass estimation procedure in the same manner as before. When multiple *Z*s are provided to `combine_estimates()` and `combine_variances()` they are treated as groupings and all other parameters will be aligned with them. When creating an object for spawning area *A*, care must be taken to only have unique `Time`/`Region` groupings. Using an object from `Estimate_P0()` with multiple *Z*s can erroneously add duplicates if you're not careful. This will cause error messages.

```{r}
# Make sure that Area has unique Time/Region breakdowns (Only Region used in this example)
Area <- data.frame(Region = unique(Multi_Z_P0_results$Region))

Area$A <- c(2822201039, #NGSV
            1884895030, # NSG
            2285649237, # SGSV
            3239913469) # SSG


Multi_Z_adult_pars <- combine_estimates(P0 = Multi_Z_P0_results, R = sex_ratio_results, S = Spawn_results, A = Area, W_F = F_W_results)

Multi_Z_adult_vars <- combine_variances(P0 = Multi_Z_P0_results, R = sex_ratio_results, S = Spawn_results, W_F = F_W_results)
```

A new object with *Z* as a grouping will now be produced which can be passed to the biomass estimation function.

```{r}
Multi_Z_adult_pars

# Estimate Biomass
Snapper_biomass_results <- Estimate_DEPM_Biomass(adult.pars = Multi_Z_adult_pars,
                                          adult.vars = Multi_Z_adult_vars)


# Create new objects
Snapper_biomass <- Snapper_biomass_results$Biomass
Snapper_Nfem <- Snapper_biomass_results$Nfem
```

These results can then be examined to determine how biomass is influenced by the specification of *Z*.

```{r fig.height=6, fig.width=6}


Snapper_biomass$Z <- as.factor(Snapper_biomass$Z) # convert Z to a factor for easy plotting

ggplot(Snapper_biomass, aes(Z,Biomass, fill = Z,
                            ymin = Biomass - SD,
                            ymax = Biomass + SD))+
  geom_crossbar( alpha = .8)+
  facet_wrap(~Region, ncol =2)+
  scale_fill_viridis_d()+
  expand_limits(y = 0)+
  scale_y_continuous(name = "Biomass (kg)", labels = scales::comma, breaks = scales::pretty_breaks(5))+
  theme_bw()

```


# References
McGarvey, R., Steer, M. A., Smart, J. J., Matthews, D. J. and Matthews, J. M.(in review). Generalizing the Parker model equation of DEPM:    incorporating size dependence of population numbers and batch fecundity

McGarvey, R., Steer, M.A., Matthews, J.M., Ward, T.M. (2018). A stage-based estimator of daily egg production. ICES J. Mar. Sci. 75(5), 1638-1646.




